{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Models for detecting BHBH, BHNS and NSNS binaries with LISA

### 2021-11-23 paper/tex/sections/comparisons.tex

\paragraph{\citet{Lau+2020}} focussed on the number of Galactic NSNS binaries that could be detected by LISA. Their study uses the same population synthesis code, COMPAS, as this work, though an earlier version. Despite this, their study finds a much larger number of detections. They make several different physical assumptions in their population synthesis, using the \citet{Fryer+2012} \textit{rapid} remnant mass prescription, limiting the maximum neutron star mass to $2 \unit{M_{\odot}}$ and not implementing PISN. However, we note that none of these assumptions strongly affect the NSNS LISA detection rate (see bottom panel of Fig.~\ref{fig:detection_rates}, models \modRapid{}, \modNSLow{} and \modNoPISN{}) and so this is unlikely to entirely account for the differences. It is also important to highlight that COMPAS has received several improvements and bug fixes since \citet{Vigna-Gomez+2018} (which contains the simulations used by \citet{Lau+2020}) and these could possibly have affected the formation rate of NSNSs.
%
Yet it is most likely that the remaining difference between our results is due to way in which we simulate the Milky Way. \citet{Lau+2020} use a model for the Milky Way similar to that of \citet{Breivik+2020}, which we use to estimate the impact of the choice of MW model in Appendix~\ref{app:mw_changes}. The Milky Way model by \cite{Breivik+2020} applies only two metallicity bins, while we consider a range of metallicities between $10^{-4}$ and $0.03$. When applying the simpler model for the Milky Way, we find that the NSNS detection rate is increased by at least a factor of two. \citet{Lau+2020} only uses a single metallicity, and so assuming that all star formation happens at a single high metallicity (which has a high efficiency of producing NSNS), could lead to an even greater overestimate of the detection rate.  Hence, we expect the low number of metallicity bins in their Milky Way model to be the main driver behind the discrepancy between our results.

@@

\paragraph{\citet{Shao+2021}} most recently investigated the detectability binaries containing BHs in LISA using BSE and a relatively simple model for the Milky Way (assuming a uniform flat disc, constant star formation and a single metallicity). They assume that kicks for NSs formed through ECSN are slightly higher than our work ($50 \unit{km}{s^{-1}}$ instead of $30 \unit{km}{s^{-1}}$). This may account for their particularly low BHNS rate (as the binaries would be more likely to disrupt), which is a factor of 20 lower than ours, but we expect their assumption of the optimistic CE scenario, reduced Wolf-Rayet winds and lower SNR detection threshold would offset this. As we show in Appendix~\ref{app:mw_changes}, their use of a simpler Milky Way model, especially with only a single metallicity, would lead to an underestimate of the BHBH and BHNS rates, which may explain the discrepancy in our results.\\

>>>>

\paragraph{\citet{Lau+2020}} focussed on the number of Galactic NSNS binaries that could be detected by LISA. Their study uses the same population synthesis code, COMPAS, as this work, though an earlier version. Despite this, their study finds a much larger number of detections. They make several different physical assumptions in their population synthesis, using the \citet{Fryer+2012} \textit{rapid} remnant mass prescription, assuming the optimistic CE scenario, limiting the maximum neutron star mass to $2 \unit{M_{\odot}}$ and not implementing PISN. However, we note that none of these assumptions strongly affect the NSNS LISA detection rate (see bottom panel of Fig.~\ref{fig:detection_rates}, models \modRapid{}, \modOpt{}, \modNSLow{} and \modNoPISN{}) and so this is unlikely to entirely account for the differences. It is also important to highlight that COMPAS has received several improvements and bug fixes since \citet{Vigna-Gomez+2018} (which contains the simulations used by \citet{Lau+2020}) and these could possibly have affected the formation rate of NSNSs.
%
Yet it is most likely that the remaining difference between our results is due to way in which we simulate the Milky Way. \citet{Lau+2020} use a model for the Milky Way similar to that of \citet{Breivik+2020}, which we use to estimate the impact of the choice of MW model in Appendix~\ref{app:mw_changes}. The Milky Way model by \cite{Breivik+2020} applies only two metallicity bins, while we consider a range of metallicities between $10^{-4}$ and $0.03$. When applying the simpler model for the Milky Way, we find that the NSNS detection rate is increased by at least a factor of two. \citet{Lau+2020} only uses a single metallicity, and so assuming that all star formation happens at a single high metallicity (which has a high efficiency of producing NSNS), could lead to an even greater overestimate of the detection rate.  Hence, we expect the low number of metallicity bins in their Milky Way model to be the main driver behind the discrepancy between our results.

@@

\paragraph{\citet{Shao+2021}} most recently investigated the detectability binaries containing BHs in LISA using BSE and a relatively simple model for the Milky Way (assuming a uniform flat disc, constant star formation and a single metallicity). They assume that kicks for NSs formed through ECSN are slightly higher than our work ($50 \unit{km}{s^{-1}}$ instead of $30 \unit{km}{s^{-1}}$). This may account for their particularly low BHNS rate (as the binaries would be more likely to disrupt), which is a factor of 20 lower than ours, but we expect their assumption of the optimistic CE scenario, reduced Wolf-Rayet winds and lower SNR detection threshold could partially offset this. As we show in Appendix~\ref{app:mw_changes}, their use of a simpler Milky Way model, especially with only a single metallicity, would lead to an underestimate of the BHBH and BHNS rates, which may explain the discrepancy in our results.\\

### 2021-11-23 paper/tex/sections/conclusion.tex

We provide predictions for the detection rate and population properties of LISA detectable BHBH, BHNS and NSNS.
%
We use a novel empirically-informed analytical model for the metallicity dependent star formation history of the Milky Way, calibrated against the APOGEE stellar spectroscopic survey. We use this to model Monte-Carlo realisations of the present-day BHBH, BHNS and NSNS populations in our Milky Way. 
%
For the binary population, we use the results of a large grid of simulations performed with the rapid population synthesis code COMPAS. These simulations have been optimised with the adaptive sampling algorithm STROOPWAFEL to preferentially sample NS and BH binaries. In total these comprise over two billion massive binaries that span 20 physics variations, which represent the most common uncertainties in binary physics.
%
To determine the detectability of sources with LISA we use the LEGWORK package \citep{Wagg+2021}, that was specifically developed for this purpose and is publicly available. 
%
We investigate the results expected for a 4- and 10-year LISA mission. Our main conclusions can be summarised as:
\begin{enumerate}
    \item \textbf{Total detections:} We predict 30-300 detections in a 4-year LISA mission, across all our simulations for varying physics assumptions. This increases to about 50-500 for a 10-year LISA mission. Although the number of detections per type can vary by about 2 orders of magnitude, we find that the total detection rate is fairly robust, among the variations we have considered (see Table~\ref{tab:detection_rates}).

@@

    \item \textbf{Probing the black hole mass distribution and the lower mass gap:} We expect LISA to predominantly detect lower mass BHs (with 90\% of BHBH and BHNSs having BH masses lower than $11 \unit{M_\odot}$ in our fiducial simulations) in stark contrast to current ground-based detectors which are heavily biased towards high mass systems. This implies that LISA can potentially make important contributions to the debate about the existence of a lower mass gap (see Fig.~\ref{fig:lower_mass_gap_variation}).

    \item \textbf{Eccentricity distribution:} We find that for all DCO types a large fraction of detectable systems still have nonzero eccentricities ($e = 0.01$) when entering the LISA band, which can be used to distinguish them from the more numerous WDWD binaries, which are largely expected to be circular. In particular, for our fiducial model, we find that this is the case for around three quarters of detectable binaries. Furthermore, around 16\% of detectable binaries have eccentricities that are so high ($e > 0.3$) that the emission at frequencies corresponding to higher order harmonics start to dominate (see Fig.~\ref{fig:fiducial_pdf_distributions}).

@@

    \item \textbf{Prospects for finding EM counterparts:} We expect about 13\% of detections with a sky localisation better than 1 degree. This fraction remains the same for a 10-year LISA mission, meaning that the number increases proportionally.
    This will be of interest for electromagnetic searches for counterparts, in particular for radio pulsar searches with SKA (see Fig.~\ref{sec:pulsar_matching}).

   \item \textbf{Benefits of extending the LISA mission:} The number of detections scale approximately as $T_{\rm obs}^{0.5}$, where $T_{\rm obs}$ is the mission length. Therefore, extending the LISA mission from 4- to 10-years increases the number of detections by about 60\%.
   A further important benefit is the improvement of the characterisation of the sources, since the relative error on the frequency derivative (which dominates the relative error in the chirp mass) scales as $T_{\rm obs}^{-2.5}$ for stationary sources (Eq.~\ref{eq:f_orb_dot_unc}).
   We find that the number of systems with chirp masses that can be measured better than 10\% increases by a factor of 2.4.
   In addition, the number of systems with a sky localisation better than one degree increases by a factor of 1.5.
   Overall, the number of sources that can be unambiguously distinguished from WDWDs increases by almost a factor of 2 (see Section~\ref{sec:WDWD_distinguish}).
\end{enumerate}

>>>>

We provide predictions for the detection rate and population properties of LISA detectable BHBH, BHNS and NSNS.
%
We use a novel empirically-informed analytical model for the metallicity dependent star formation history of the Milky Way, calibrated against the APOGEE stellar spectroscopic survey. We use this to model Monte-Carlo realisations of the present-day BHBH, BHNS and NSNS populations in our Milky Way. 
%
For the binary population, we use the results of a large grid of simulations performed with the rapid population synthesis code COMPAS \citefloorp{}. These simulations have been optimised with the adaptive sampling algorithm STROOPWAFEL to preferentially sample NS and BH binaries. In total these comprise over two billion massive binaries that span 20 physics variations, which represent the most common uncertainties in binary physics.
%
To determine the detectability of sources with LISA we use the LEGWORK package \citep{Wagg+2021}, that was specifically developed for this purpose and is publicly available. 
%
We investigate the results expected for a 4- and 10-year LISA mission. Our main conclusions can be summarised as:
\begin{enumerate}
    \item \textbf{Total detections:} We predict 30-300 detections in a 4-year LISA mission, across all our simulations for varying physics assumptions. This increases to about 50-500 for a 10-year LISA mission. Although the number of detections per type can vary by about 2 orders of magnitude, we find that the total detection rate is fairly robust, among the variations we have considered (see Table~\ref{tab:detection_rates}).

@@

    \item \textbf{Probing the black hole mass distribution and the lower mass gap:} We expect LISA to predominantly detect lower mass BHs (with 90\% of BHBH and BHNSs having BH masses lower than $11 \unit{M_\odot}$ in our fiducial simulations) in stark contrast to current ground-based detectors which are heavily biased towards high mass systems. For our fiducial simulation, we predict that approximately 69 systems with a component with a mass between $2.5$-$5 \unit{M_\odot}$ would be detected by a 4-year LISA. This implies that LISA can potentially make important contributions to the debate about the existence of a lower mass gap (see Fig.~\ref{fig:lower_mass_gap_variation}).

    \item \textbf{Eccentricity distribution:} We find that for all DCO types a large fraction of detectable systems still have nonzero eccentricities ($e > 0.01$) when entering the LISA band, which can be used to distinguish them from the more numerous WDWD binaries, which are largely expected to be circular. In particular, for our fiducial model, we find that this is the case for around three quarters of detectable binaries. Furthermore, around 16\% of detectable binaries have eccentricities that are so high ($e > 0.3$) that the emission at frequencies corresponding to higher order harmonics start to dominate (see Fig.~\ref{fig:fiducial_pdf_distributions}).

@@

    \item \textbf{Prospects for finding EM counterparts:} We expect about 13\% of detections with a sky localisation better than 1 degree for our fiducial model (though the fraction remains roughly constant among model variations). This fraction remains the same for a 10-year LISA mission, meaning that the number increases proportionally. This will be of interest for electromagnetic searches for counterparts, in particular for radio pulsar searches with SKA (see Fig.~\ref{sec:pulsar_matching}).

   \item \textbf{Benefits of extending the LISA mission:} The number of detections scale approximately as $\sqrt{T_{\rm obs}}$, where $T_{\rm obs}$ is the mission length. Therefore, extending the LISA mission from 4- to 10-years increases the number of detections by about 60\% for each model variation. A further important benefit is the improvement of the characterisation of the sources, since the relative error on the frequency derivative (which dominates the relative error in the chirp mass) scales as $T_{\rm obs}^{-2.5}$ for stationary sources (Eq.~\ref{eq:f_orb_dot_unc}). We find that the number of systems with chirp masses that can be measured better than 10\% increases by a factor of 2.4 for each model variation.
   In addition, the number of systems with a sky localisation better than one degree increases by a factor of 1.5. Overall, the number of sources that can be unambiguously distinguished from WDWDs increases by almost a factor of 2 (see Section~\ref{sec:WDWD_distinguish}).
\end{enumerate}

### 2021-11-23 paper/tex/sections/introduction.tex

Earlier work on BHBHs, BHNSs and NSNSs in LISA has used a variety of population synthesis codes, Milky Way models and LISA specifications, resulting in a wide range of predictions \citep{Nelemans+2001,Liu+2009,Belczynski+2010,Liu+2014,Lamberts+2019,Lau+2020,Breivik+2020,Sesana+2020}. We build upon previous efforts but with several important improvements. We explore the effects of varying binary physics assumptions by repeating our analysis for \nModels{} different models and comparing the effects on the detection rate and distributions of source parameters. We use a model for the Milky Way that accounts for the chemical enrichment history and is calibrated on the APOGEE survey \citep{Majewski+2017,Frankel+2018}, whereas most others did not consider the effect of metallicity in detail (see however \citealp{Lamberts+2019, Sesana+2020}). We provide a detailed treatment of the eccentricity of detectable sources, both for the inspiral evolution as well as gravitational wave signal during the LISA mission. Moreover, our binary population synthesis simulation is the most extensive of its kind to date and makes use of the adaptive sampling algorithm STROOPWAFEL \citep{Broekgaarden+2019, Broekgaarden+2021}. Overall we simulate over 2 billion massive binaries to produce the DCO populations used in this work. We find that this large number of simulations is important to reduce the sampling noise even when using adaptive importance sampling.

>>>>

Earlier work on BHBHs, BHNSs and NSNSs in LISA has used a variety of population synthesis codes, Milky Way models and LISA specifications, resulting in a wide range of predictions \citep{Nelemans+2001,Liu+2009,Belczynski+2010,Liu+2014,Lamberts+2019,Lau+2020,Breivik+2020,Sesana+2020, Shao+2021}. We build upon previous efforts but with several important improvements. We explore the effects of varying binary physics assumptions by repeating our analysis for \nModels{} different models and comparing the effects on the detection rate and distributions of source parameters \citefloorp{}. We use a model for the Milky Way that accounts for the chemical enrichment history and is calibrated on the APOGEE survey \citep{Majewski+2017,Frankel+2018}, whereas most others did not consider the effect of metallicity in detail (see however \citealp{Lamberts+2019, Sesana+2020}). We provide a detailed treatment of the eccentricity of detectable sources, both for the inspiral evolution as well as gravitational wave signal during the LISA mission. Moreover, our binary population synthesis simulation is the most extensive of its kind to date and makes use of the adaptive sampling algorithm STROOPWAFEL \citep{Broekgaarden+2019, Broekgaarden+2021}. Overall we simulate over 2 billion massive binaries to produce the DCO populations used in this work. We find that this large number of simulations is important to reduce the sampling noise even when using adaptive importance sampling.

### 2021-11-23 paper/tex/sections/methods.tex

We use the grid of \nModels{} binary population synthesis simulations recently presented in \citet{Broekgaarden+2021} and Broekgaarden et al.\ (in preparation). This grid of simulations is synthesised using the rapid population synthesis code \href{https://compas.science}{COMPAS} \citep{Stevenson+2017,Vigna-Gomez+2018,Stevenson+2019,Broekgaarden+2019}. COMPAS follows the approach of the population synthesis code BSE \citep{Hurley+2000,Hurley+2002} and uses fitting formula and rapid algorithms to efficiently predict the final fate of binary systems. The code is open source and documented in the papers listed above, the online documentation\footnote{\url{https://compas.science}} and in the methods paper \citep{COMPAS:2021methodsPaper}. We summarise the main assumptions and settings relevant for this work in Appendix~\ref{app:pop_synth}.

The result of the simulations is a sample of binaries, which, for each metallicity $Z$, have $N_{\rm binaries}$ binaries with parameters
\begin{equation}
    \mathbf{b}_{{Z, i}} = \{m_1, m_2, a_{\rm DCO}, e_{\rm DCO}, t_{\rm evolve}, t_{\rm inspiral}, w\},
\end{equation}
for $i = 1, 2, \dots, N_{\rm binaries}$, where $m_1$ and $m_2$ are the primary and secondary masses, $a_{\rm DCO}$ and $e_{\rm DCO}$ are the semi-major axis and eccentricity at the moment of double compact object (DCO) formation, $t_{\rm evolve}$ is the time between the binary's zero-age main sequence and DCO formation, $t_{\rm inspiral}$ is the time between DCO formation (that is immediately after the second supernova in the system) and gravitational-wave merger and $w$ is the adaptive importance sampling weight assigned by STROOPWAFEL \cite[][Eq.~7]{Broekgaarden+2019}. We sample from these sets of parameters when creating synthetic galaxies.

@@

We match each set of galaxy parameters $\mathbf{g}_{{j}}$, to a random set of binary parameters $\mathbf{b}_{{Z, i}}$, by drawing a binary from the closest metallicity bin to the metallicity in $\mathbf{g}_{{j}}$.

@@

We use \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} to calculate the signal-to-noise ratio for each binary and the package ensures that enough harmonics are computed for each binary such that the error on the gravitational-wave luminosity remains below 1\%.

@@

We then determine the fraction of binaries that are detectable in each Milky Way instance by summing the adaptive importance sampling weights of the binaries that have an SNR greater than 7, and dividing by the total weights in the simulation. We multiply this fraction by the total number in the Milky Way to find a detection rate (which we write as a total number of detections per LISA mission)
\begin{equation}
    N_{\rm detect} = \frac{\sum_{i = 0}^{N_{\rm MW}} w_i \phi(i)}{\sum_{i = 0}^{N_{\rm MW}} w_i} N_{\rm DCO},
\end{equation}
where $\phi(i) = 1$ if a binary is detectable and $0$ otherwise. We calculate the detection rate by Monte Carlo sampling 2500 Milky Way instances (each containing 200,000 DCOs) for each DCO type and every physics variation in order to obtain values for the uncertainty on the expected detection rate.

>>>>

We use the grid of \nModels{} binary population synthesis simulations recently presented in \citet{Broekgaarden+2021} and Broekgaarden et al.\ (in prep.). This grid of simulations is synthesised using the rapid population synthesis code \href{https://compas.science}{COMPAS} \citep{Stevenson+2017,Vigna-Gomez+2018,Stevenson+2019,Broekgaarden+2019}. COMPAS follows the approach of the population synthesis code BSE \citep{Hurley+2000,Hurley+2002} and uses fitting formula and rapid algorithms to efficiently predict the final fate of binary systems. The code is open source and documented in the papers listed above, the online documentation\footnote{\url{https://compas.science}} and in the methods paper \citep{COMPAS:2021methodsPaper}. We summarise the main assumptions and settings relevant for this work in Appendix~\ref{app:pop_synth}.

The result of the simulations is a sample of binaries, which, for each metallicity $Z$, have $N_{\rm binaries}$ binaries with parameters
\begin{equation}
    \mathbf{b}_{{Z, i}} = \{m_1, m_2, a_{\rm DCO}, e_{\rm DCO}, t_{\rm evolve}, t_{\rm inspiral}, w\},
\end{equation}
for $i = 1, 2, \dots, N_{\rm binaries}$, where $m_1$ and $m_2$ are the primary and secondary masses, $a_{\rm DCO}$ and $e_{\rm DCO}$ are the semi-major axis and eccentricity at the moment of double compact object (DCO) formation, $t_{\rm evolve}$ is the time between the binary's zero-age main sequence and DCO formation, $t_{\rm inspiral}$ is the time between DCO formation (that is, immediately after the second supernova in the system) and gravitational-wave merger and $w$ is the adaptive importance sampling weight assigned by STROOPWAFEL \cite[][Eq.~7]{Broekgaarden+2019}. We sample from these sets of parameters when creating synthetic galaxies.

@@

We match each set of galaxy parameters $\mathbf{g}_{{j}}$, to a random set of binary parameters $\mathbf{b}_{{Z, i}}$, by drawing a binary from the closest metallicity bin to the metallicity in $\mathbf{g}_{{j}}$. If the metallicity in $\mathbf{g}_{{j}}$ is below the minimum COMPAS metallicity bin ($Z = 10^{-4}$), we use this minimum bin. If the metallicity in $\mathbf{g}_{{j}}$ is above the maximum COMPAS metallicity bin ($Z = 0.03$), we use a randomly selected bin from the five highest metallicity bins.

@@

We use \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} \citep{Wagg+2021} to calculate the signal-to-noise ratio for each binary and the package ensures that enough harmonics are computed for each binary such that the error on the gravitational-wave luminosity remains below 1\%.

@@

We then determine the fraction of binaries that are detectable in each Milky Way instance by summing the adaptive importance sampling weights of the binaries that have an SNR greater than 7, and dividing by the total weights in the simulation. We multiply this fraction by $N_{\rm DCO}$ to find a detection rate (which we write as a total number of detections per LISA mission)
\begin{equation}
    N_{\rm detect} = \frac{\sum_{i = 0}^{N_{\rm MW}} w_i \phi(i)}{\sum_{i = 0}^{N_{\rm MW}} w_i} N_{\rm DCO},
\end{equation}
where $\phi(i) = 1$ if a binary is detectable and $0$ otherwise. We calculate the detection rate by Monte Carlo sampling 2500 Milky Way instances (each containing 200,000 DCOs) for each DCO type and every physics variation in order to obtain values for the uncertainty on the expected detection rate.
